---
title: 'Lab 3: Individual 1'
author: 'John Chandara \<chandara@iastate.edu\>'
date: '2/19/20'
output: html_document
---

```{r}
library(tidyverse)
library(ggtern)

dat <- readxl::read_xls('GSS.xls')
```

# GSS Data Exploration

## Data Cleaning

```{r}
indx <- sapply(dat, is.character)
dat[indx] <- lapply(dat[indx], function(x) as.factor(as.character(x)))
names(dat) <- c('year', 'happiness', 'party', 'residence', 'sex', 'education', 'children', 'marriage', 'id', 'ballot')
dat <- droplevels(dat[dat$happiness != 'No answer' & dat$happiness != 'Don\'t know' & dat$happiness != 'Not applicable', ])
dat <- droplevels(dat[dat$party != 'No answer', ])
dat$happiness <- factor(dat$happiness, c('Very happy', 'Pretty happy', 'Not too happy'))
```

```{r}

rotatedAxisElementText = function(angle,position='x'){
  angle     = angle[1]; 
  position  = position[1]
  positions = list(x=0,y=90,top=180,right=270)
  if(!position %in% names(positions))
    stop(sprintf("'position' must be one of [%s]",paste(names(positions),collapse=", ")),call.=FALSE)
  if(!is.numeric(angle))
    stop("'angle' must be numeric",call.=FALSE)
  rads  = (angle - positions[[ position ]])*pi/180
  hjust = 0.5*(1 - sin(rads))
  vjust = 0.5*(1 + cos(rads))
  element_text(angle=angle,vjust=vjust,hjust=hjust)
}
```

## Exploration

### How does the happiness of a respondent relate to the political party affiliation?

```{r}
  dat2 <- dat

  parties <- levels(dat2$party)
  levels(dat2$party)[levels(dat2$party) == 'Not str democrat' | levels(dat2$party) == 'Strong democrat'] <- 'Democrat'

  levels(dat2$party)[levels(dat2$party) == 'Not str republican' | levels(dat2$party) == 'Strong republican'] <- 'Republican'

  levels(dat2$party)[levels(dat2$party) == 'Don\'t know' | levels(dat2$party) == 'Other party'] <- 'Republican'

  levels(dat2$party)[levels(dat2$party) == 'Independent' | levels(dat2$party) == 'Ind,near dem' | levels(dat2$party) == 'Ind,near rep'] <- 'Independent'

  dat2 %>%
    ggplot(aes(x = party, fill = happiness)) + geom_bar(position = 'fill')
```
  Based on the results, republicans appear to be much more happier than independent and democratic respondents by a well margin.

### Is political affiliation affected by education?

```{r}
  dat2$education <- factor(dat2$education, c('No answer', 'Don\'t know','Lt high school', 'High school','Junior college', 'Bachelor', 'Graduate'))

  filtered <- droplevels(dat2[dat2$education != 'No answer' & dat2$education != 'Don\'t know', ])

  filtered %>%
    ggplot(aes(x = education, fill = education)) + geom_bar() + facet_wrap(~party) + theme(axis.text.x = rotatedAxisElementText(90,'x'))

```

From this data, we can conclude that respondent with the highest education level in junior college typically swing twords independent or democratic ideologies. Time afterwards, their ideology may swing a bit republican however, this isn't strong enough as a small percentage more of full graduates are democrats.

  

### How does the political affiliation affect place of residence?


```{r}
  levels(dat2$residence)[levels(dat2$residence) == 'E. nor. central' | levels(dat2$residence) == 'W. nor. central'] <- 'North'

  levels(dat2$residence)[levels(dat2$residence) == 'New england' | levels(dat2$residence) == 'Middle atlantic' | levels(dat2$residence) == 'South atlantic'] <- 'East'

  levels(dat2$residence)[levels(dat2$residence) == 'E. sou. central' | levels(dat2$residence) == 'W. sou. central'] <- 'South'

  levels(dat2$residence)[levels(dat2$residence) == 'Mountain' | levels(dat2$residence) == 'Pacific' | levels(dat2$residence) == 'Ind,near rep'] <- 'West'

  dat2$residence <- factor(dat2$residence, c('North', 'East', 'South', 'West', 'Foreign'))

  dat2 %>%
    ggplot(aes(x = residence, fill = party)) + geom_bar(position = 'fill', width = 1) + coord_polar()
```

For the most part, all regions of the United States appear to be in quite equal in representation. Though, we see a bit more Republican respondents in the North and West. For democrats, we see a magnitude more in the south. As for Foreign respondent, there appears to be less republican-centered respondent.
